---
title: "_MarZICM_ for calculating marginal mediation effects within zero-inflated compositional mediators"
author:
- name: Quran Wu
  affiliation: University of Florida
- name: Zhigang Li
  affiliation: University of Florida
  email: zhigang.li@ufl.edu
package: MarZICM
output:
  BiocStyle::html_document
vignette: |
  %\VignetteIndexEntry{MarZICM for calculating marginal mediation effects within zero-inflated compositional mediators}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r style, echo=FALSE, results='asis'}
BiocStyle::markdown()
```


# Introduction
MarZICM provides a way of calculating mediation effects within zero-inflated compositional structure mediators marginally. It can be applied in microbiome studies with taxa as mediator. The inputs could be either relative abundance (RA), or absolute abundance (AA), which will be transformed to RA in analysis. A library size, or sequencing depth is required for analysis. MarZICM will assume the taxon follows either an univariate beta distribution, or a mixed beta distribution, to improve the flexibility and robustness to model misspecifications. 

The marginal outcome model on taxon $M_j$ is: 
$$
Y=\beta_0+\beta_1M_j+\beta_21_{M_j>0}+\beta_3X+\beta_4X1_{M_j>0}+\beta_5XM_j+\epsilon
$$
The mean of RA of $M_j$ is:
$$
\log(\frac{\mu_j}{1-\mu_j})=\alpha_0 + \alpha_1X
$$
The probability of observing structure zero on $M_j$ is: 
$$
\log(\frac{\Delta_j}{1-\Delta_j})=\gamma_0 + \gamma_1X
$$

The Natural Indirect Effect (NIE) will be automatically decomposed into NIE$_1$ and NIE$_2$. 

# Input for MarZICM() function

- `Experiment_dat` A SummarizedExperiment object containing microbiome data as assay and covariates, outcome and library size as colData. The microbiome data could be relative abundance or absolute abundance. Missing value should be imputed in advance.
- `lib_name`      Name of library size variable within colData.
- `y_name`   Name of outcome variable within colData.
- `x_name` Name of covariate of interest within colData.
- `conf_name` Name of confounders within colData. Defaule is NULL, meaning no confounder.
- `mediator_mix_range` Number of mixtures in mediator. Default is 1, meaning no mixture.
- `transfer_to_RA` Logical variable indicating whether the microbiome data should be transferred to relative abundance. Default is FALSE. If TRUE, microbiome data will be rescaled by its row sum.
- `num_cores` Number of CPU cores to be used in parallelization task.
- `adjust_method` P value adjustment method. Same as p.adjust. Default is "fdr".
- `fdr_rate` FDR cutoff for significance. Default is 0.2.
- `taxDropThresh` The threshold of dropping taxon due to high zero percentage. Default is 0.9, meaning taxon will be dropped for analysis if zero percentage is higher than 90\%.
- `taxDropCount` The threshold of dropping taxon due to not enough non-zero observation counts. Default is 20, meaning taxon will be dropped if non-zero observation is less than 20.
- `SDThresh` The threshold of dropping taxon due to low coefficient of variation (CV) to avoid constant taxon. Default is 0.05, meaning any taxon has CV less than 0.05 will be dropped.
- `SDx` The threshold of stopping analysis due to low CV of covariate of interest. Default is 0.05, meaning when CV of covariate of interest is less than 0.05, the analysis will be stopped.
- `SDy` The threshold of stopping analysis due to low CV of outcome. Default is 0.05, meaning when CV of outcome. is less than 0.05, the analysis will be stopped.

# Output for IFAA() function

A `list` of `4` datasets containing the results for `NIE1`, `NIE2`, `NDE`, and `NIE`.
Each dataset has row representing each taxon, 6 columns for `Estimates`, `Standard Error`,
`Lower bound for 95% Confidence Interval`, `Upper bound for 95% Confidence Interval`,
`Adjusted p value`, `Significance indicator`.

# Example
```{r}
library(MarZICM)
library(SummarizedExperiment)
library(dirmult)

## A make up example with 1 taxon and 100 subjects.
set.seed(1)
nSub <- 200
nTaxa <- 10
## generate covariate of interest X
X <- rbinom(nSub, 1, 0.5)
## generate mean of each taxon. All taxon are having the same mean for simplicity.
mu <- exp(-5 + X) / (1 + exp(-5 + X))
phi <- 10

## generate true RA
M_taxon<-t(sapply(mu,function(x) rdirichlet(n=1,rep(x*phi,nTaxa))))

P_zero <- exp(-3 + 0.3 * X) / (1 + exp(-3 + 0.3 * X))

non_zero_ind <- t(sapply(P_zero,function(x) 1-rbinom(nTaxa,1,rep(x,nTaxa))))

True_RA<-t(apply(M_taxon*non_zero_ind,1,function(x) x/sum(x)))

## generate outcome Y based on true RA
Y <- 1 + 100 * True_RA[,1] + 5 * (True_RA[,1] > 0) + X + rnorm(nSub)

## library size was set to 10,000 for all subjects for simplicity.
libsize <- 10000

## generate observed RA
observed_AA <- floor(M_taxon*libsize*non_zero_ind)

observed_RA <- t(apply(observed_AA,1,function(x) x/sum(x)))
colnames(observed_RA)<-paste0("rawCount",seq_len(nTaxa))
## Construct SummerizedExperiment object
CovData <- cbind(Y = Y, X = X, libsize = libsize)
test_dat <-
  SummarizedExperiment(assays = list(MicrobData = t(observed_RA)), colData = CovData)

```
Above steps could be ignored if a SummarizedExperiment object is already available. 
```{r}
## run the analysis
res <- MarZICM(
  Experiment_dat = test_dat,
  lib_name = "libsize",
  y_name = "Y",
  x_name = "X",
  num_cores = 1,
  mediator_mix_range = 1
)


```

The output result containing 4 datasets for NIE$_1$, NIE$_2$, NDE, NIE, respectively. The NIE$_1$, for example, could be extracted by:
```{r}
NIE1 <- res$NIE1_save
```
The significant result could be extracted by: 
```{r}
subset(NIE1, significance == TRUE)
```
From the result, we can find that `rawCount1` is an significant mediator of mediating the effect of $X$ on $Y$. 


```{r sessionInfo, echo=FALSE}
sessionInfo()
```
