---
title: "_MarZIC_ for calculating marginal mediation effects within zero-inflated compositional mediators"
author:
- name: Quran Wu
  affiliation: University of Florida
- name: Zhigang Li
  affiliation: University of Florida
  email: zhigang.li@ufl.edu
package: MarZIC
output:
  BiocStyle::html_document
vignette: |
  %\VignetteIndexEntry{MarZIC for calculating marginal mediation effects within zero-inflated compositional mediators}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r style, echo=FALSE, results='asis'}
BiocStyle::markdown()
```


# Introduction
MarZIC estimates and tests the marginal mediation effects of zero-inflated compositional mediators. It can be applied to microbiome studies to estimate and test the mediation effects of relative abundances (RA). MarZIC decomposes the mediation effect into two components of which one is the mediation effect induced by the continuous change on the numeric scale of the mediator and the other is the mediation effect induced by the binary change from zero to non-zero status of the mediator. For a microbiome data, let $M_j$ denote the RA of the $j$th taxon/OTU/ASV. And let $Y$, $X$ and $Z$ denote the outcome, the independent variable and confounder variable respectively, where $Z$ could contain multiple variables. The marginal outcome model for $Y$ is: 
$$
Y=\beta_0+\beta_1M_j+\beta_21_{M_j>0}+\beta_3X+\beta_4X1_{M_j>0}+\beta_5XM_j+\beta_6Z+\epsilon
$$
The mean of $M_j$ depend on $X$ and $Z$ through the following equation:
$$
\log(\frac{\mu_j}{1-\mu_j})=\alpha_0 + \alpha_1X+\alpha_2Z
$$
Let $\Delta_j=P(M_j=0)$ which is the probability of observing a structural zero (ie, true zero). We use the following logistic model for this probability:  
$$
\log(\frac{\Delta_j}{1-\Delta_j})=\gamma_0 + \gamma_1X+\gamma_2Z
$$

# Input for MarZIC() function

- `Experiment_dat` A SummarizedExperiment object containing microbiome data, covariates (ie, independent variable and confounders), outcome and library size. The microbiome data could be relative abundance or absolute abundance. If it is absolute abundance, it will be converted into relative abundance in the analysis by default.
- `lib_name`      Name of library size variable.
- `y_name`   Name of outcome variable
- `x_name` Name of covariate of interest
- `conf_name` Name of confounders Defaule is NULL, meaning no confounder.
- `taxa_of_interest` A character vector for taxa names indicating taxa that should be analyzed. Default is NULL, meaning all taxa should be included into analysis. 
- `mediator_mix_range` Number of mixtures in mediator. Default is 1, meaning no mixture.
- `transfer_to_RA` Logical variable indicating whether the microbiome data should be transferred into relative abundance. Default is TRUE. If FALSE, microbiome data will not be transferred.
- `num_cores` Number of CPU cores to be used in parallelization task.
- `adjust_method` P value adjustment method. Same as p.adjust in R. Default is "fdr".
- `fdr_rate` FDR cutoff for significance. Default is 0.05.
- `taxDropThresh` The threshold of dropping taxon due to high zero percentage. Default is 0.9, meaning taxon will be dropped for analysis if zero percentage is higher than 90\%.
- `taxDropCount` The threshold of dropping taxon due to not enough non-zero observation counts. Default is 4 * (length(conf_name)+2), meaning taxon will be dropped if non-zero observation is less than four times of the total number of covariates (including the independent variable) plus intercept. 
- `zero_prop_NIE2` The threshold of zero percentage for calculating NIE2. Default is 0.1, meaning NIE2 will be calculated for taxon with zero percentage greater than 10\%.
- `zero_count_NIE2` The threshold of zero counts for calculating NIE2. Default is 4 * (length(conf_name)+2), meaning NIE2 will be calculated for taxon with zero counts greater than four times of number of covariates plus 1.
- `SDThresh` The threshold of dropping taxon due to low coefficient of variation (CV) to avoid constant taxon. Default is 0.05, meaning any taxon has CV less than 0.05 will be dropped.
- `SDx` The threshold of stopping analysis due to low CV of covariate of interest. Default is 0.05, meaning when CV of covariate of interest is less than 0.05, the analysis will be stopped.
- `SDy` The threshold of stopping analysis due to low CV of outcome. Default is 0.05, meaning when CV of outcome. is less than 0.05, the analysis will be stopped.

# Output for IFAA() function

A `list` of `4` datasets containing the results for `NIE1`, `NIE2`, `NDE`, and `NIE`.
Each dataset has row representing each taxon, 6 columns for `Estimates`, `Standard Error`,
`Lower bound for 95% Confidence Interval`, `Upper bound for 95% Confidence Interval`,
`Adjusted p value`, `Significance indicator`.

# Example
```{r}
library(MarZIC)
library(SummarizedExperiment)
library(dirmult)

## A make up example with 1 taxon and 100 subjects.
set.seed(1)
nSub <- 200
nTaxa <- 10
## generate covariate of interest X
X <- rbinom(nSub, 1, 0.5)
## generate confounders
conf1<-rnorm(nSub)
conf2<-rbinom(nSub,1,0.5)

## generate mean of each taxon. All taxon are having the same mean for simplicity.
mu <- exp(-5 + X + 0.1 * conf1 + 0.1 * conf2) / 
  (1 + exp(-5 + X + 0.1 * conf1 + 0.1 * conf2))
phi <- 10

## generate true RA
M_taxon<-t(sapply(mu,function(x) rdirichlet(n=1,rep(x*phi,nTaxa))))

P_zero <- exp(-3 + 0.3 * X + 0.1 * conf1 + 0.1 * conf2) / 
  (1 + exp(-3 + 0.3 * X + 0.1 * conf1 + 0.1 * conf2))

non_zero_ind <- t(sapply(P_zero,function(x) 1-rbinom(nTaxa,1,rep(x,nTaxa))))

True_RA<-t(apply(M_taxon*non_zero_ind,1,function(x) x/sum(x)))

## generate outcome Y based on true RA
Y <- 1 + 100 * True_RA[,1] + 5 * (True_RA[,1] > 0) + X + conf1 + conf2 + rnorm(nSub)

## library size was set to 10,000 for all subjects for simplicity.
libsize <- 10000

## generate observed RA
observed_AA <- floor(M_taxon*libsize*non_zero_ind)

observed_RA <- t(apply(observed_AA,1,function(x) x/sum(x)))
colnames(observed_RA)<-paste0("rawCount",seq_len(nTaxa))
## Construct SummerizedExperiment object
CovData <- cbind(Y = Y, X = X, libsize = libsize, conf1 = conf1, conf2 = conf2)
test_dat <-
  SummarizedExperiment(assays = list(MicrobData = t(observed_RA)), colData = CovData)

```
Above steps could be ignored if a SummarizedExperiment object is already available. 
Suppose we're interested in the mediation effect in `rawCount1`, `rawCount2`, and `rawCount3`, the analysis could be done as: 
```{r}
## run the analysis
res <- MarZIC(
  Experiment_dat = test_dat,
  lib_name = "libsize",
  y_name = "Y",
  x_name = "X",
  conf_name = c("conf1","conf2"),
  taxa_of_interest = c("rawCount1","rawCount2","rawCount3"),
  num_cores = 1,
  mediator_mix_range = 1
)


```
The Natural Indirect Effect (NIE) will be decomposed into NIE$_1$ and NIE$_2$. The output result containing 4 datasets for NIE$_1$, NIE$_2$, NDE, NIE, respectively. The NIE$_1$, for example, could be extracted by:
```{r}
NIE1 <- res$NIE1_save
```
The significant result could be extracted by: 
```{r}
subset(NIE1, significance == TRUE)
```
From the result, we can find that `rawCount1` is an significant mediator of mediating the effect of $X$ on $Y$. 


```{r sessionInfo, echo=FALSE}
sessionInfo()
```
